Quotients of the magmatic operad:\\ lattice structures and convergent rewrite systems

### 2018-09-09 CAs_operads.tex
Relecture partie 3 (avant partie 3.3.2)

In order to transform the rewrite relation induced by~\eqref{equ:rew_1}
into a convergent one, we apply the Buchberger algorithm for
operads~\cite[Section 3.7]{DK10} with respect to the lexicographic order
on prefix words. We first focus on the special case $\gamma = 3$.
\medbreak

@@

\begin{Theorem} \label{thm:convergent_rewrite_rule_CAs_3}
    The set $\Rew$ of rewrite rules containing~\eqref{equ:rew_1},
    and \eqref{equ:rew_2}---\eqref{equ:rew_11} is a finite
    convergent presentation of~$\CAs{3}$.
\end{Theorem}
\begin{proof}
    Let us show that the rewrite relation $\RewContext$ induced by
    $\Rew$ is convergent. First, for every relation $\Tfr \Rew \Tfr'$,
    we have $\Tfr > \Tfr'$. Therefore, by
    Lemma~\ref{lem:prefix_word_termination}, $\RewContext$ is
    terminating. Moreover, the greatest degree of a tree appearing in
    $\Rew$ is~$7$ so that, from Lemma~\ref{lem:degree_confluence}, to
    show that $\RewContext$ is convergent, it is enough to prove that
    each tree of degree at most $13$ admits exactly one normal form.
    Equivalently, this amounts to show that the number of normal forms
    of trees of arity $n\leq 14$ is equal to $\#\CAs{3}(n)$. By computer
    exploration, we get the same sequence
    \begin{equation} \label{equ:dimensions_CAs_3}
        1, 1, 2, 4, 8, 14, 20, 19, 16, 14, 14, 15, 16, 17
    \end{equation}
    for $\#\CAs{3}(n)$ and for the numbers of normal forms of arity $n$,
    when $ 1 \leq n \leq 14$, which proves
    Theorem~\ref{thm:convergent_rewrite_rule_CAs_3}.
\end{proof}
\medbreak

@@

\begin{Proposition} \label{prop:PBW_basis_CAs_3}
    The set of the trees avoiding as subtrees the ones appearing as
    left members of $\Rew$ is a Poincaré-Birkhoff-Witt basis
    of~$\CAs{3}$.
\end{Proposition}
\begin{proof}
    By definition of PBW bases and
    Theorem~\ref{thm:convergent_rewrite_rule_CAs_3}, the set
    $\NormalForms_{\RewContext}$ is a PBW basis of $\CAs{3}$ where
    $\RewContext$ is the rewrite relation induced by $\Rew$. Now, by
    Lemma~\ref{lem:normal_forms_avoiding}, $\NormalForms_{\RewContext}$
    can be described as the set of the trees avoiding the left members
    of~$\Rew$, whence the statement.
\end{proof}
\medbreak

@@

Let us describe the normal forms for arity $n\geq 11$. For that, we
consider the following "zig-zag" of degree $k \geq 0$

@@

  z_k := \Product \circ_1 ( \Product \circ_2 ( \Product
  \circ_1 ( \Product \circ_2 \cdots
  \Product ) \cdots ).

@@

For instance, $z_4$ and $z_5$ are the following trees

@@

@@

@@

The normal forms are decomposed into two families.
The first family is the set of the $n-1$ trees
$\left(z_k \circ_{k+1} z_{n-1-k}\right)_{1 \leq k \leq n-1}$.
For example,
\begin{equation}\label{equ:normal_forms_type_A}
   z_8 \circ_{9} z_{3} =

@@

      \node[NodeST](11)at(11.00,-11.11){\begin{math}\Product\end{math}};
      \node[NodeST](13)at(13.00,-5.56){\begin{math}\Product\end{math}};

@@

      \node[NodeST](17)at(17.00,-5.56){\begin{math}\Product\end{math}};
      \node[NodeST](21)at(21.00,-8.33){\begin{math}\Product\end{math}};
      \node[NodeST](23)at(23.00,-2.78){\begin{math}\Product\end{math}};

@@

      \node(r)at(15.00,2.08){};

@@

The second family is composed of the following four trees

@@

  \end{tikzpicture}, \qquad

@@

Indeed, each of these $n+3$ trees avoids the left membres of the

@@

\begin{Proposition}
  The family~\eqref{equ:normal_forms_type_B} is absorbing for each
  partial composition.
  \end{Proposition}

\begin{proof}
  None of the trees of type~\eqref{equ:normal_forms_type_A}
  contains the left or the right member of the congruence relation
  $\CongrCAs{3}$, so that they are alone in their equivalence classes.
  Moreover, the trees of type~\eqref{equ:normal_forms_type_B}
  contain the right member of $\CongrCAs{3}$, so that the composition of
  such a tree with anothter tree is not alone in its
  equivalence class, and thus it does not belong to the
  family~\eqref{equ:normal_forms_type_A}.
\end{proof}

By computer explorations, we are able to conjecture the multiplication
tables of each partial compositions. However, the obtained tables are not
simply described. For instance, the partial composition
$\Tfr\circ_i\Tfr'$ with $\Tfr$ of type~\eqref{equ:normal_forms_type_A}
and $\Tfr'$ of type~\eqref{equ:normal_forms_type_B} is described by $36$
cases depending on the values of various parameters associated to $\Tfr$,
$\Tfr'$ and $i$.

\medbreak
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsubsection{Higher comb associative operads}
\label{sec:higher_comb_associative_operads}

>>>>

In order to transform the rewrite relation induced by~\eqref{equ:rew_1}
into a convergent one, we apply the Buchberger algorithm for
operads~\cite[Section 3.7]{DK10} with respect to the lexicographic order
on prefix words. We first focus on the special case~$\gamma = 3$.
\medbreak

@@

\begin{Theorem} \label{thm:convergent_rewrite_rule_CAs_3}
    The set $\Rew$ of rewrite rules containing~\eqref{equ:rew_1},
    and \eqref{equ:rew_2}---\eqref{equ:rew_11} is a finite
    convergent presentation of~$\CAs{3}$.
\end{Theorem}
\begin{proof}
    Let us show that the rewrite relation $\RewContext$ induced by
    $\Rew$ is convergent. First, for every relation $\Tfr \Rew \Tfr'$,
    we have $\Tfr > \Tfr'$. Therefore, by
    Lemma~\ref{lem:prefix_word_termination}, $\RewContext$ is
    terminating. Moreover, the greatest degree of a tree appearing in
    $\Rew$ is~$7$ so that, from Lemma~\ref{lem:degree_confluence}, to
    show that $\RewContext$ is convergent, it is enough to prove that
    each tree of degree at most $13$ admits exactly one normal form.
    Equivalently, this amounts to show that the number of normal forms
    of trees of arity $n\leq 14$ is equal to $\#\CAs{3}(n)$. By computer
    exploration, we get the same sequence
    \begin{equation} \label{equ:dimensions_CAs_3}
        1, 1, 2, 4, 8, 14, 20, 19, 16, 14, 14, 15, 16, 17
    \end{equation}
    for $\#\CAs{3}(n)$ and for the numbers of normal forms of arity $n$,
    when $ 1 \leq n \leq 14$, which proves the statement of the
    theorem.
\end{proof}
\medbreak

@@

\begin{Proposition} \label{prop:PBW_basis_CAs_3}
    The set of the trees avoiding as subtrees the ones appearing as
    left members of $\Rew$ is a PBW basis of~$\CAs{3}$.
\end{Proposition}
\begin{proof}
    By definition of PBW bases and
    Theorem~\ref{thm:convergent_rewrite_rule_CAs_3}, the set
    $\NormalForms_{\RewContext}$ is a PBW basis of $\CAs{3}$ where
    $\RewContext$ is the rewrite relation induced by $\Rew$. Now, by
    Lemma~\ref{lem:normal_forms_avoiding}, $\NormalForms_{\RewContext}$
    can be described as the set of the trees avoiding the left members
    of~$\Rew$, whence the statement.
\end{proof}
\medbreak

@@

Let us describe the elements of the PBW basis of $\CAs{3}$ for arity
$n \geq 11$. By Proposition~\ref{prop:PBW_basis_CAs_3}, these elements
are the normal forms of the rewrite relation induced by $\Rew$. Let for
any $d \geq 0$, the binary tree $\Zfr_d$ defined recursively by

@@

    \Zfr_d :=
    \begin{cases}
        \Leaf & \mbox{if } d = 0, \\
        \Zfr_{d - 1}
        \circ_{\left\lfloor \frac{d}{2} \right\rfloor + 1} \Product
        & \mbox{otherwise}.
    \end{cases}

@@

For instance,

@@

    \Zfr_4 =

@@

    \Zfr_5 =

@@

The normal forms split into two families. The first one is the set
of the $n - 1$ trees of the form
\begin{equation} \label{equ:normal_forms_type_A}
    \Zfr_d \circ_{d + 1} \Zfr_{n - 1 - d},
for any $d \in [n - 1]$. For example, for $n = 12$,
    z_8 \circ_9 z_3 =

@@

        \node[NodeST](11)at(11.00,-11.11)
        \node[NodeST](13)at(13.00,-5.56)

@@

        \node[NodeST](17)at(17.00,-5.56)
        \node[NodeST](21)at(21.00,-8.33)
        \node[NodeST](23)at(23.00,-2.78)

@@

        \node(r)at(15.00,2.25){};

@@

is a tree of this first family. The second family contains the
following four trees

@@

    \end{tikzpicture}\,, \qquad

@@

Observe that each of these $n + 4$ trees avoids the left members of the

@@

\begin{Proposition} \label{prop:absorbing_normal_forms_type_B}
    For any two trees $\Tfr$ and $\Tfr'$ of the
    form~\eqref{equ:normal_forms_type_B} and $i \in [|\Tfr|]$, the
    normal form for the rewrite relation induced by $\Rew$ of
    $\Tfr \circ_i \Tfr'$ is a tree of the
    form~\eqref{equ:normal_forms_type_B}.
\end{Proposition}
\begin{proof}
    None of the trees of the form~\eqref{equ:normal_forms_type_A}
    contain the left or the right member of the rewrite
    rule~\eqref{equ:rew_1} generating the congruence relation
    $\CongrCAs{3}$. Hence, these trees are alone in their equivalence
    classes. Moreover, the trees of the
    form~\eqref{equ:normal_forms_type_B} contain the right member of
    the rewrite rule~\eqref{equ:rew_1} generating the congruence
    relation $\CongrCAs{3}$, so that the composition of such a tree
    with another tree is not alone in its equivalence class, and thus
    it does not belong to the family~\eqref{equ:normal_forms_type_A}.
\end{proof}
\medbreak

Proposition~\ref{prop:absorbing_normal_forms_type_B} says that the
family of trees~\eqref{equ:normal_forms_type_B} is absorbing
for the partial composition.
\medbreak

Computer explorations allow us to conjecture the multiplication table
of the exhibited PBW basis of $\CAs{3}$. However, we do note have a
simple description of this table. For instance, the partial composition
$\Tfr\circ_i\Tfr'$ where $\Tfr$ is a tree of the
form~\eqref{equ:normal_forms_type_A} and $\Tfr'$ is a tree of the
form~\eqref{equ:normal_forms_type_B} can be fully described by $36$
cases depending on the values of various parameters associated with
$\Tfr$, $\Tfr'$, and~$i$.
\medbreak

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsubsection{Higher comb associative operads}
\label{sec:higher_comb_associative_operads}
